---
title: "Hello R Markdown"
author: "Frida Gomam"
date: 2015-07-23T21:13:14-05:00
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(tidyverse)
library(DT)
library(plotly)
```

# R Markdown

# War, what is it good for (Dec 15th, 2017)  

[Rahul Oka](http://anthropology.nd.edu/faculty-and-staff/faculty-by-alpha/rahul-oka/), [Agustin Fuentes](http://anthropology.nd.edu/faculty-and-staff/faculty-by-alpha/agustin-fuentes/), [Susan Sheridan](http://anthropology.nd.edu/faculty-and-staff/faculty-by-alpha/susan-sheridan/), [Mark Golitko](http://anthropology.nd.edu/faculty-and-staff/faculty-by-alpha/mark-golitko/), [Nam Kim](https://www.anthropology.wisc.edu/staff/kim-nam-c/), and I recently published a paper in PNAS looking at how demographics affects both the size of a war group and conflict causalities. 

 __Here is the gist__: 

       
Many scholars have noted that in hunter-gather populations, the number of people involved in fighting can be as much as 40% of the population, but this percentage drops in big state-level societies. One of the first people to note this was Larry Keeley in his War Before Civilization, first published in 1994. He noted that the percentage of male war deaths was higher in nonstates than in states. Others, such as Steven Pinker, have taken these data to mean that “civilization” has a pacifying element. As Pinker puts it “violence has declined over long stretches of time, and today we may be living in the most peaceable era in our species’  existence” (Pinker 2011: xxi). Using data from Keeley and other scholars, he concludes that before the emergence of the state, the violent death rate was around 15%.

Raul Oka wondered if this could be due to population size. In other words, is this drop in the number of people involved in war (what we call the “war group”) due to the effects of living in a state or is it a reflection of broader issues that appear when population increases?
We collected data on population and war group size from 295 societies and on war group size and conflict-related casualties from 430 historical conflicts going back to 2500 B.C. We show that, as population goes up, the proportion of people involved in conflict decreases.  In other words, it is not that states prevent violence, but rather that other factors, such as the need to feed, cloth, and arm a war group makes it harder to involve a large percentage of people in more complex societies. 

__You can read [the paper here](http://www.pnas.org/content/early/2017/12/07/1713972114.short?rss=1)__

**press coverage**

* Michael Price [covered our paper in Science](http://www.sciencemag.org/news/2017/12/why-human-society-isn-t-more-or-less-violent-past), where Steven Pinker has some thoughts to share...

* My university was nice enough to [cover it as well](https://today.appstate.edu/2017/12/13/war-group)

* [Inverse interview with Rahul Oka](https://www.inverse.com/article/39263-war-groups-are-humans-history-more-less-violent)



* It also got some international attention in [Italian]( https://www.galileonet.it/2017/12/violenza-civilizzazione-guerre/) and [Russian!]( https://www.nkj.ru/news/32809/)



##Looking at the data from our paper
This is a working version looking at some of the data from Oka et al. to try to explain it and use some visualization techniques to examine the results interactively


```{r datatable for war, warning=FALSE, message= FALSE, echo=FALSE}
war <- read_csv("PopSizeTemp.csv") ##### or temp file?
```


The following figure is similar to the published figure, but you can hover over each point to examine the data more carefully


```{r figure 1a1, warning = FALSE, echo = FALSE}
#plot_Fig1a1 <- ggplot() + geom_point(data = war, aes(x= log(Population), y = log(Overall_War_Group_Size_W), color = Time_Type))+ geom_smooth(data = war, aes(x= log(Population), y = log(Overall_War_Group_Size_W), group=1),method="lm") + geom_point(data= war, aes(x= log(Population), y = log(W.P), color = Time_Type)) + geom_smooth(data = war, aes(x= log(Population), y = log(W.P), group=1),method="lm")


plot_Fig1a1 <- ggplot() + geom_point(data = war, aes(x= log(Population), y = log(Overall_War_Group_Size_W), color = Time_Type, text = paste("Society: ", Society) ))+ geom_smooth(data = war, aes(x= log(Population), y = log(Overall_War_Group_Size_W), group=1),method="lm") + geom_point(data= war, aes(x= log(Population), y = log(W.P), color = Time_Type, text = paste("Society: ", Society))) + geom_smooth(data = war, aes(x= log(Population), y = log(W.P), group=1),method="lm")

```


```{r, warning=FALSE, message= FALSE, echo=FALSE}
ggplotly(plot_Fig1a1, height = 600, width = 600)
```
#Fig1a: Scaled distribution of P vs. W and W/P from Dataset S1 (n = 295). W and W/p







# Including Plots

You can also embed plots. See Figure \@ref(fig:pie) for example:

```{r pie, fig.cap='A fancy pie chart.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```
