---
title: Human Evolution
author: Marc
date: '2018-07-20'
slug: human-evolution
categories: []
tags: []
header:
  caption: ''
  image: ''
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(DT)
library(plotly)
```






#Exploring hominin data

This is a temp version of data for my Human Evolution course next Fall. Wanted to spend the first day showing some data and talking about how we can visualize hominin evolution.


##interactive plot

Below is a scatterplot of the cranial capacity of various hominin species (species designation taken from referenced sources. Your mileage may vary as to whether these are true species or not...). Age is in millions of years. Mouse over to see info (update coming with more interactive features). 

```{r, message=FALSE, error=FALSE, warning=FALSE, echo=FALSE}
cc <-read_csv("CC_data_Updated_2.csv") #read in data
 p2 <- ggplot(cc, aes(x = Age_in_MA_rounded, y = Cranial_capacity, color = Species_2, label = Specimen)) + geom_point() 
#p2
```



```{r, message=FALSE, error=FALSE, warning=FALSE, echo=FALSE}
ggplotly(p2)

```


##boxplot

Boxplot of the cranial capacities:



```{r, message=FALSE, error=FALSE, warning=FALSE, echo=FALSE}
p_box <-  ggplot(cc, aes(x = reorder(Species_2, Cranial_capacity, na.rm = TRUE) , y = Cranial_capacity, fill = Species_2)) + geom_boxplot() + coord_flip() +labs(y= "Cranial Capacity", x = NULL)

p_box

#ggplotly(p_box, height = 600, width = 800)
```



##Ridge plot


The graph below is a [ridge plot](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html), which is a density plot aligned on a horizontal scale. 


```{r ridge_plots_2, message=FALSE, error=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=8 }
cc <-read_csv("CC_data_Updated_4.csv") #read in data

cc$Specimen<-iconv(enc2utf8(cc$Specimen),sub="byte") #fix the format of Specimen col

#cc %>% group_by(Species_2) %>% summarise(max(Age_in_MA_rounded), min(Age_in_MA_rounded))



plot10 <- ggplot(cc, aes(x = Cranial_capacity, y = reorder(Species_2, -date2 ),fill = Species_2)) + geom_density_ridges() + theme_ridges() + labs( caption = "data from Schoenemann 2013 & Hawks et al 2017", fill ="Species")

plot10a <- plot10 + annotate("text", label = "", x =1, y =1) + theme(axis.title.y = element_blank()) + labs(x="Cranial Capacity in cm^3")

plot10a

```


##some data

Tables of data (still trying to make this look nice...)

```{r datatable}
cc %>% group_by(Species_2) %>% summarise(mean(Cranial_capacity))



```






